---
title: "R Utilities Functions"
author: "Akil Elkamel"
date: "6/11/2020"
output:
  beamer_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()
```

## pivot_longer() - Definition

The pivot_longer() function from the tidyr package, "lengthens" data, increasing the number of rows and decreasing the number of columns. The inverse transformation is pivot_wider()

**Usage:**

```{r eval=FALSE}
pivot_longer(
  data,
  cols,
  names_to = "name",
  names_prefix = NULL,
  names_sep = NULL,
  names_pattern = NULL,
  names_ptypes = list(),
  names_transform = list(),
  names_repair = "check_unique",
  values_to = "value",
  values_drop_na = FALSE,
  values_ptypes = list(),
  values_transform = list(),
  ...
)
```


## pivot_longer() - Example 1
```{r echo=FALSE, message=FALSE}
library(tidyverse)
```


```{r message=FALSE}
# Simplest case where column names are character data
relig_income
```


## pivot_longer() - Example 1

```{r message=FALSE}
relig_income %>%
  pivot_longer(-religion, names_to = "income", values_to = "count")
```

## pivot_longer() - Example 2

```{r}
# Slightly more complex case where columns have common prefix,
# and missing missings are structural so should be dropped.
billboard
```

## pivot_longer() - Example 2

```{r}
# Slightly more complex case where columns have common prefix,
# and missing missings are structural so should be dropped.
billboard
```

## pivot_longer() - Example 2

```{r}
billboard %>%
 pivot_longer(
   cols = starts_with("wk"),
   names_to = "week",
   names_prefix = "wk",
   values_to = "rank",
   values_drop_na = TRUE
 )
```


## pivot_longer() - Example 3

```{r}
# Multiple variables stored in column names
who
```

## pivot_longer() - Example 3

```{r}
who %>% pivot_longer(
  cols = new_sp_m014:newrel_f65,
  names_to = c("diagnosis", "gender", "age"),
  names_pattern = "new_?(.*)_(.)(.*)",
  values_to = "count"
)
```


## pivot_longer() - Example 4

```{r}
# Multiple observations per row
anscombe
```

## pivot_longer() - Example 4

```{r}
anscombe %>%
 pivot_longer(everything(),
   names_to = c(".value", "set"),
   names_pattern = "(.)(.)"
 )
```